---
title: "Met500 Plots"
output: html_notebook
---
```{r libs,eval=TRUE, include=FALSE}
library(tidyverse)
library(DESeq2)
library(purrr)
library(ggplot2)
library(dplyr)
library(RColorBrewer)
library(pheatmap)
library(SummarizedExperiment)
library(knitr)
library(reshape2)
library(limma)
library(edgeR)
library(PCAtools)
library(clusterProfiler)
library(org.Hs.eg.db)
```

#loading in rsd and formatting SE object
```{r data, eval=TRUE, include=FALSE}
gse2 <- readRDS(file = "/Users/U1006746/Desktop/counts.AK.output/dds_met_gene_counts.rds")
list <- c()
```

```{r filter, eval=TRUE, include=FALSE}
#filter out genes (must have more than 2 counts) >= 3???
keepIt <- rowSums(cpm(gse2) > 10) >= 300
sumexp <- gse2[keepIt,]

gene_info <- as.data.frame(rowData(sumexp)@listData)
assay_info <- as.data.frame(assay(sumexp))
sample_info <- as.data.frame(colData(sumexp)@listData)

#filtered out 1000 (Not assessed for age)
sample_info <- sample_info%>%
  filter(Age != 1000)

#updating SE for ages we want to keep
ages <- sample_info$Run
sumexp <- sumexp[ ,ages]

sample_info <- as.data.frame(colData(sumexp)@listData)
gene_info <- as.data.frame(rowData(sumexp)@listData)
assay_info <- as.data.frame(assay(sumexp))
gene_info$external_gene_name <- ifelse(is.na(gene_info$external_gene_name), gene_info$GENEID, gene_info$external_gene_name)
gene_info$external_gene_name <- ifelse(gene_info$external_gene_name == "", gene_info$GENEID, gene_info$external_gene_name)
sample_info <- sample_info[ ,-c(3:30)]
sample_info <- sample_info[ ,-c(4:8)]
sample_info <- sample_info[ ,-c(18:19)]

#format dataset to lower and make new body column for all randomness
sample_info <- sample_info %>%
  dplyr::mutate(body_site = tolower(body_site),
                PRIMARY_TUMOR_LOCATION = tolower(PRIMARY_TUMOR_LOCATION))

sample_info$BODY_SITE <- str_extract(sample_info$body_site, "^[^\\s_]+")
sample_info$PRIMARY_TUMOR_LOCATION <- str_extract(sample_info$PRIMARY_TUMOR_LOCATION, "^[^\\s_]+")
sample_info$BODY_SITE <- ifelse(sample_info$BODY_SITE == "abdomen", "abdominal", sample_info$BODY_SITE)
sample_info$PRIMARY_TUMOR_LOCATION <- ifelse(sample_info$PRIMARY_TUMOR_LOCATION == "abdomen", "abdominal", sample_info$PRIMARY_TUMOR_LOCATION)

#updating sumexp object 
sumexp$BODY_SITE <- sample_info$BODY_SITE
sumexp$PRIMARY_TUMOR_LOCATION <- sample_info$PRIMARY_TUMOR_LOCATION

#table data 
age <- sumexp$Age
patient <- sumexp$submitted_subject_id
body_site <- sumexp$BODY_SITE
gender <- sumexp$Gender
tumor <- sumexp$PRIMARY_METASTATIC_TUMOR
prim_location <- sumexp$PRIMARY_TUMOR_LOCATION
table(patient, gender, body_site, age, tumor)

```

```{r}
sumexp3 <- sumexp
assay(sumexp3) <- log2(cpm(assay(sumexp3)) + 1)
assaySumE3_info <- as.data.frame(assay(sumexp3))
```

```{r}
#rid of dups
plottingMET_data <- data.frame(subject = sumexp3$submitted_subject_id,
                           sample = sumexp3$Run,
                           body = sumexp3$BODY_SITE,
                           gene_expression = colSums(assays(sumexp3)$counts),
                           tumor = sumexp3$PRIMARY_METASTATIC_TUMOR,
                           prim_loc = sumexp3$PRIMARY_TUMOR_LOCATION,
                           age = sumexp3$Age,
                           sex = sumexp3$Gender,
                           hist = sumexp3$HISTOLOGICAL_TYPE)

plottingMETFilt <- plottingMET_data[!duplicated(plottingMET_data[c("subject", "body","tumor","prim_loc","age","sex","hist")]), ]
rows <- rownames(plottingMETFilt)

plottingMETFilt$prim_loc <- factor(plottingMETFilt$prim_loc)

#updatin no dups for sumexp
sumexp <- sumexp[ ,rows]
sample_info <- as.data.frame(colData(sumexp)@listData)
```

```{r}
library(tidySummarizedExperiment)
sumexp3 <- sumexp3 %>%
  group_by(PRIMARY_TUMOR_LOCATION)

sumexp3<- sumexp3 %>%
  filter(.sample %in% rows)

sumexp3$.sample <- factor(sumexp3$.sample, levels = unique(sumexp3$.sample[order(sumexp3$Gender)]))
sumexp3$.sample <- factor(sumexp3$.sample, levels = unique(sumexp3$.sample[order(sumexp3$PRIMARY_TUMOR_LOCATION)]))

```

Summary of Data:
* 1145 samples, not paired 
* 1070 Metastatic sites and 71 Primary sites
* 1132 subjects
* focus on just Metastatic sites


```{r}
ggthemr("light")
plottingMETFilt %>%
  filter(!tumor %in% c("NA","Primary"),
         body %in% c("bone", "brain", "abdominal", "chest","pelvic", "liver")) %>%
  ggplot(aes(x = body)) +
  geom_bar()+
  xlab("Tumor Site") +
  ylab("Number of Samples") +
  ggtitle("Number of Metastatic Tumor Samples per Body Site") 


ggsave('barplot_liver_.png', width = 8, height = 8)
```
**Fig 1a. Looking at the number of samples per metastatic tumor sites, these were picked as total samples over 20. I chose to investigate liver metastatic tumors it has the highest number of samples.**


```{r}
ggthemr("dust")
plottingMETFilt %>%
  filter(!tumor %in% c("NA","Primary"),
         body == "liver",
         prim_loc %in% c("bile", "esophagus", "prostate", "breast", "lung")) %>%
  ggplot(aes(x = prim_loc)) +
  geom_bar(aes(fill = sex))+
  xlab("Primary Tumor Site (of Liver Met)") +
  ylab("Number of Samples") +
  ggtitle("Number of Liver Metastatic Samples per Primary Tumor Body Site")

ggsave('barplotGENDEr_liver_.png', width = 8, height = 8)

```
**Fig 1b. Investigates how gender is distributed across two primary tumor locations for liver metastasized tumors. There is similar distribution of males in females in each primary tumor location for esophagus and bile and similar sample sizes so I picked these to investigate.**

```{r}

ggthemr("grape")
sumexp3$.sample <- factor(sumexp3$.sample, levels = unique(sumexp3$.sample[order(sumexp3$PRIMARY_TUMOR_LOCATION)]))


sumexp3 %>%
filter(BODY_SITE == "liver",
       PRIMARY_TUMOR_LOCATION %in% c("bile", "esophagus"),
       PRIMARY_METASTATIC_TUMOR == "Metastatic") %>%
 ggplot(aes(x = .sample, y = counts, fill = PRIMARY_TUMOR_LOCATION)) +
  geom_boxplot()+
  ggtitle("Distribution of Gene Expression per Sample ")+
  xlab("Sample") +
  ylab("Gene Expression (log, CPM)") +
  labs(fill = "Primary Tumor") +
  theme(legend.text = element_text(size = 6))+
  coord_flip() 

ggsave('boxplot_liver_GE.png', width = 8, height = 10)

```

**Fig 2. Looking at gene expression in counts per million to account for size and its distribution for each of the liver metastatic body site samples. Filled by primary tumor location either bile ducts or esophagus. We see ever so slightly esophagus have a little higher median gene expression, is it significant? **

```{r}

ggthemr("light")
plottingMETFilt %>%
  filter(!tumor %in% c("NA","Primary"),
         body == "liver",
         prim_loc %in% c("bile", "esophagus")) %>%
  ggplot(aes(x = prim_loc, y = gene_expression)) +
  geom_boxplot(position = "dodge") +
  labs(x = "Gender", y = "Total Gene Expression (log/CPM)", color = "Tumor Type") +
  ggtitle("Total Gene Expression Across Primary Tumor Locations")+
  geom_text(data = plottingMETFilt %>% 
              filter(!tumor %in% c("NA","Primary"),body == "liver",prim_loc %in% c("bile", "esophagus")) %>%
              group_by(prim_loc) %>% 
              summarise(top = max(gene_expression), 
                        n=n()),
            aes(x= prim_loc, y=0, label= paste0("n = ", n)), 
            size = 3.5, fontface = "bold", vjust = -2
             )

ggsave('boxplotTOTAL_liver_.png', width = 8, height = 8)

```
**Fig 3. Total gene expression in counts per million (to account for size) and log transformed (for visualization purposes) for all liver metastatic tumors samples. It is comparing the expression between from bile vs esophagus primary tumors and esophagus is slightly higher. They do have comparable sample sizes (31 total samples for this subset).**


```{r}
#subset just no duplicates abdonimal
samp_info_absMET <- sample_info %>%
  filter(BODY_SITE == "liver", 
         PRIMARY_METASTATIC_TUMOR == "Metastatic",
         PRIMARY_TUMOR_LOCATION %in% c("bile", "esophagus"))
keep <- samp_info_absMET$Run
sumexpAbs <- sumexp[ ,keep]

#dds is filtered for this subset question with only primary tumor locations bein bile and esophagus 
assays(sumexpAbs)$counts <- round(assays(sumexpAbs)$counts)
dds2 <- DESeqDataSet(sumexpAbs, design = ~ PRIMARY_TUMOR_LOCATION)
dds2$PRIMARY_TUMOR_LOCATION <- factor(dds2$PRIMARY_TUMOR_LOCATION, levels = c("bile","esophagus"))

vsd2 <- vst(dds2, blind = TRUE)
p2 <- pca(assay(vsd2), metadata = colData(vsd2), removeVar = 0.1)
biplot(p2, colby = "PRIMARY_TUMOR_LOCATION", lab = NULL, legendPosition = 'right', legendTitleSize = 8, legendLabSize = 8, title = "PCA of Bile vs Esophagus Tumors" )
ggsave('metpcas.png', width = 8, height = 5)
```
**Fig 4. The PCA shows some distinction between primary tumor locations, however there is some mixing.**

```{r}
#DESeq DE
#using medain-of-ratio method to normalize
#estimate size factor for dds
dds2 = estimateSizeFactors(dds2)

#estimating dispersion
dds2 = estimateDispersions(dds2)

#test
dds2 = nbinomWaldTest(dds2)
contrast1 <- c("PRIMARY_TUMOR_LOCATION", "bile", "esophagus")
resDESeq2 = results(dds2, contrast = contrast1)

#ordering by significant p-values
resDESeq2 <- resDESeq2[order(resDESeq2$padj),]
head(resDESeq2)

#summary of <0.05 p-value
summary(resDESeq2, alpha = 0.05)


#just sig genes
padj.cutoff <- 0.05
sigRes4 <- resDESeq2 %>%
        data.frame() %>%
        filter(padj < padj.cutoff) %>%
        rownames_to_column(var = "gene")
sigRes4 <- left_join(sigRes4, gene_info, by = c("gene" = "GENEID"))
rownames(sigRes4) <- sigRes4$gene
sigRes4 <- sigRes4[,-1]


#results and gene names no filter
sigRes_all <- resDESeq2 %>%
        data.frame() %>%
        rownames_to_column(var = "gene")
sigRes_all <- left_join(sigRes_all, gene_info, by = c("gene" = "GENEID"))
rownames(sigRes_all) <- sigRes_all$gene
sigRes_all <- sigRes_all[,-1]

#filter for heatmap
# sigRes_fil2 <- sigRes_all[(sigRes_all$baseMean > 150) & abs(sigRes_all$log2FoldChange > 2),]
# summary(sigRes_fil2, alpha > 0.05)
```

```{r}
look <- sigRes_all %>%
 filter(!grepl("^ENSG", external_gene_name),
        padj <= 0.05,
        log2FoldChange <= -1)
```

```{r}
select = c("CDCA5","HMGA1", "BRI3BP", "UBE2S", "KIAA0903", "TSPAN5", "FGFR4", "GK5","GASK1A", "ADPRM","MED12L","MYH11","ACTG2", "PRDM8", "MED12L", "ZNF594", "PRDM8")


select_cool <- c("GK5", "UBE2S", "CDCA5", "BIRC5", "TK1")
ggthemr("grape")

EnhancedVolcano(toptable = sigRes_all,
                x = "log2FoldChange",
                y = "padj",
                lab = sigRes_all$external_gene_name,
                select = select_cool,
                boxedLabels = TRUE,
                drawConnectors = TRUE,
                widthConnectors = 1.0,
                colConnectors = 'black',
                pCutoff = .05,
                FCcutoff = 1)



ggsave('Met_volcano_plot.png', width = 15, height = 10)
```
**Fig 5. Shows genes in red that both have "significant" differences between the two and have distinct unregulated or downregulated log fold changes (how much gene is expressed).**

```{r}
#getting sample info
DDS2sample_info <- as.data.frame(colData(dds2))

#getting genes
genes.kept <- sigRes4[sigRes4$external_gene_name %in% select_cool, ]
genes.kept <- rownames(genes.kept)

#getting counts/treatment/gene names situated
count.table <- counts(dds2, normalized=T)[genes.kept, ]
cols <- colnames(count.table)
DDS2sample_info <- DDS2sample_info %>%
  filter(rownames(DDS2sample_info) %in% cols)
colnames(count.table) = paste(DDS2sample_info$PRIMARY_TUMOR_LOCATION, colnames(count.table), sep = " : ")

#gene id names
count.table <- count.table %>%
  as.data.frame() %>%
  rownames_to_column(var = "gene")
count.table <-  left_join(count.table, gene_info, by = c("gene" = "GENEID"))
rownames(count.table) <- count.table$external_gene_name
count.table <- count.table[,-33]
count.table <- count.table[,-1]

#heatmap z scores
count.table.z <- t(apply(count.table, 1, scale))
colnames(count.table.z) <- colnames(count.table)

#colors

myCols2 <- c("bile : SRR4307033", "bile : SRR4307364","bile : SRR4306923","bile : SRR4307368","esophagus : SRR8762949","esophagus : SRR8762625","esophagus : SRR8762960","esophagus : SRR8763004")
row_idx2 <- which(colnames(count.table.z) %in% myCols2)


fontcolors <- rep('grey', ncol(count.table.z))
fontcolors[11] <- "black" #esophagus
fontcolors[12] <- "#0000FF" #bile
fontcolors[14] <- "#0000FF"  #bile
fontcolors[15] <- "#0000FF" #bile

fontcolors[23] <- "black" #esophagus
fontcolors[11] <- "#0000FF" #bile
fontcolors[27] <- "black" #esophagus
fontcolors[28] <- "black" #esophagus
fontcolors[29] <- 'black' #esophagus

 fontcolors_rows <- rep('black', nrow(count.table.z))
 fontcolors_rows[1] <- "#0000FF"
 fontcolors_rows[2] <- "#0000FF"
 fontcolors_rows[3] <- "#0000FF"
 fontcolors_rows[5] <- "black"
 fontcolors_rows[4] <- "#0000FF"



#heatmap
h4 <- Heatmap(count.table.z, cluster_rows = T, column_labels = colnames(count.table.z), name = "Z-score", row_labels = rownames(count.table.z),
              column_names_gp = gpar(col = fontcolors, fontface = "bold"),
             row_names_gp = gpar(col = fontcolors_rows, fontface = "bold"))

```

```{r printing heatmap, echo=FALSE}
png("heatmap_MET.png", res = 250, width = 4000, height = 2000)
print(h4)
dev.off()
```
**Fig 6. Tend to be specific unregulated genes in esophagus vs bile. For example CDK5 is upregulated in bile duct cancers and not in esophagus. **


#GSEA
```{r}
#ordered results table
sigResall_gse <- sigRes_all %>%
  filter(!grepl("^ENSG",external_gene_name)) %>%
  rownames_to_column(var = "geneID")

#mading it not be versions
sigResall_gse <- sigResall_gse %>%
  mutate(geneID = str_replace(geneID,"\\..*$", ""))
rownames(sigResall_gse) <- sigResall_gse$geneID
sigResall_gse <- sigResall_gse[,-1]
```

```{r}
#gene list ordered by descending LFC
gene_list2 <- sigResall_gse$log2FoldChange
names(gene_list2) <- rownames(sigResall_gse)
gene_list2 <- na.omit(gene_list2)
gene_list2 = sort(gene_list2,decreasing = TRUE)
gene_list2
```

```{r}
#running gsea
met_gse <- gseGO(gene_list2,
                    ont = "BP",
                    keyType = "ENSEMBL",
                    OrgDb = "org.Hs.eg.db",
                    eps = 1e-300)

new <- as.data.frame(met_gse)
```

```{r}
gseaplot(met_gse, geneSetID = 102, title = met_gse$Description[102])
ggsave('Met_gsea.png', width = 15, height = 10)
```

