---
title: "Met500 Plots"
output: html_notebook
---
```{r libs,eval=TRUE, include=FALSE}
library(tidyverse)
library(DESeq2)
library(purrr)
library(ggplot2)
library(dplyr)
library(RColorBrewer)
library(pheatmap)
library(SummarizedExperiment)
library(knitr)
library(reshape2)
library(limma)
library(edgeR)
library(PCAtools)
library(clusterProfiler)
library(org.Hs.eg.db)
```

#loading in rsd and formatting SE object
```{r data, eval=TRUE, include=FALSE}
gse2 <- readRDS(file = "/Users/U1006746/Desktop/counts.AK.output/dds_met_gene_counts.rds")
list <- c()
```

```{r filter, eval=TRUE, include=FALSE}
#filter out genes (must have more than 2 counts) >= 3???
keepIt <- rowSums(cpm(gse2) > 10) >= 300
sumexp <- gse2[keepIt,]

gene_info <- as.data.frame(rowData(sumexp)@listData)
assay_info <- as.data.frame(assay(sumexp))
sample_info <- as.data.frame(colData(sumexp)@listData)

#filtered out 1000 (Not assessed for age)
sample_info <- sample_info%>%
  filter(Age != 1000)

#updating SE for ages we want to keep
ages <- sample_info$Run
sumexp <- sumexp[ ,ages]

sample_info <- as.data.frame(colData(sumexp)@listData)
gene_info <- as.data.frame(rowData(sumexp)@listData)
assay_info <- as.data.frame(assay(sumexp))
gene_info$external_gene_name <- ifelse(is.na(gene_info$external_gene_name), gene_info$GENEID, gene_info$external_gene_name)
gene_info$external_gene_name <- ifelse(gene_info$external_gene_name == "", gene_info$GENEID, gene_info$external_gene_name)
sample_info <- sample_info[ ,-c(3:30)]
sample_info <- sample_info[ ,-c(4:8)]
sample_info <- sample_info[ ,-c(18:19)]

#format dataset to lower and make new body column for all randomness
sample_info <- sample_info %>%
  dplyr::mutate(body_site = tolower(body_site),
                PRIMARY_TUMOR_LOCATION = tolower(PRIMARY_TUMOR_LOCATION))

sample_info$BODY_SITE <- str_extract(sample_info$body_site, "^[^\\s_]+")
sample_info$PRIMARY_TUMOR_LOCATION <- str_extract(sample_info$PRIMARY_TUMOR_LOCATION, "^[^\\s_]+")
sample_info$BODY_SITE <- ifelse(sample_info$BODY_SITE == "abdomen", "abdominal", sample_info$BODY_SITE)
sample_info$PRIMARY_TUMOR_LOCATION <- ifelse(sample_info$PRIMARY_TUMOR_LOCATION == "abdomen", "abdominal", sample_info$PRIMARY_TUMOR_LOCATION)
#sample_info$BODY_SITE <- sample_info[complete.cases(sample_info$BODY_SITE), ]

#updating sumexp object 
sumexp$BODY_SITE <- sample_info$BODY_SITE
sumexp$PRIMARY_TUMOR_LOCATION <- sample_info$PRIMARY_TUMOR_LOCATION

#table data 
age <- sumexp$Age
patient <- sumexp$submitted_subject_id
body_site <- sumexp$BODY_SITE
gender <- sumexp$Gender
tumor <- sumexp$PRIMARY_METASTATIC_TUMOR
prim_location <- sumexp$PRIMARY_TUMOR_LOCATION
table(patient, gender, body_site, age, tumor)

```

```{r}
sumexp3 <- sumexp
assay(sumexp3) <- log2(cpm(assay(sumexp3)) + 1)
```

```{r}
#rid of dups
plottingMET_data <- data.frame(subject = sumexp3$submitted_subject_id,
                           sample = sumexp3$Run,
                           body = sumexp3$BODY_SITE,
                           gene_expression = colSums(assays(sumexp3)$counts),
                           tumor = sumexp3$PRIMARY_METASTATIC_TUMOR,
                           prim_loc = sumexp3$PRIMARY_TUMOR_LOCATION,
                           age = sumexp3$Age,
                           sex = sumexp3$Gender,
                           hist = sumexp3$HISTOLOGICAL_TYPE,
                           library = sumexp3$LibrarySelection)

#got rid of duplicate runs (all information is same in row exept total gene expression)
plottingMETFilt <- plottingMET_data[!duplicated(plottingMET_data[c("subject", "body","tumor","prim_loc","age","sex","hist")]), ]
rows <- rownames(plottingMETFilt)

plottingMETFilt <- plottingMETFilt %>%
  filter(tumor != "NA", prim_loc != "?")

plottingMETFilt$prim_loc <- factor(plottingMETFilt$prim_loc)
plottingMETFilt$age <- as.integer(plottingMETFilt$age)

#updatin no dups for sumexp
sumexp <- sumexp[ ,rows]
sample_info <- as.data.frame(colData(sumexp)@listData)
```

```{r}
library(tidySummarizedExperiment)
sumexp3 <- sumexp3 %>%
  group_by(PRIMARY_TUMOR_LOCATION)

sumexp3<- sumexp3 %>%
  filter(.sample %in% rows)

sumexp3$.sample <- factor(sumexp3$.sample, levels = unique(sumexp3$.sample[order(sumexp3$Gender)]))
sumexp3$.sample <- factor(sumexp3$.sample, levels = unique(sumexp3$.sample[order(sumexp3$PRIMARY_TUMOR_LOCATION)]))

```


Summary of Data:
* 1141 samples, not paired 
* 1070 Metastatic sites and 71 Primary sites (no matched samples)
* 1132 subjects
* focus on just Metastatic sites

```{r}

ggthemr("light")
plottingMETFilt %>%
  filter(!tumor %in% c("Primary"),
         body %in% c("bone", "brain", "abdominal", "chest","pelvic", "liver"),
         sex == "M") %>%
  ggplot(aes(x = body)) +
  geom_bar()+
  xlab("Tumor Site") +
  ylab("Number of Samples") +
  ggtitle("Number of Metastatic Tumor Samples per Body Site") 
#ggsave('barplot_liver_.png', width = 8, height = 8)

```
**Fig 1a. Looking at the number of samples per metastatic tumor sites, these were picked as total samples over 20. I chose to investigate liver metastatic tumors it has the highest number of samples.**


```{r}
ggthemr("dust")
plottingMETFilt %>%
  filter(!tumor %in% c("NA","Primary"),
         sex == "M",
         body == "liver",
         prim_loc %in% c("bile", "esophagus", "prostate", "breast", "lung")) %>%
  ggplot(aes(x = prim_loc)) +
  geom_bar()+
  xlab("Primary Tumor Location") +
  ylab("Number of Samples") +
  ggtitle("Number of Liver Metastatic Samples per Primary Tumor Location")

ggsave('barplotGENDEr_liver_.png', width = 10, height = 10)

```
**Fig 1b. Investigates how gender is distributed across two primary tumor locations for liver metastasized tumors. There is similar distribution of males in females in each primary tumor location for esophagus and bile and similar sample sizes so I picked these to investigate.**

```{r}
#ages subset of 60 and 70 y/o, liver mets
Ages <- plottingMETFilt %>%
  mutate(under = age == 60, 
          over = age == 70) %>%
  filter(under == "TRUE" | over == "TRUE",
         body == "liver",) %>%
  ggplot(aes(x = age)) +
  geom_bar()+
  xlab("") +
  ylab("") +
  theme(axis.text.x = element_text(angle = 90))+
  ggtitle("")

```

```{r}

ggthemr("grape")
sumexp3$.sample <- factor(sumexp3$.sample, levels = unique(sumexp3$.sample[order(sumexp3$PRIMARY_TUMOR_LOCATION)]))


sumexp3 %>%
filter(BODY_SITE == "liver",
      Gender == "M",
       PRIMARY_TUMOR_LOCATION %in% c("bile", "esophagus"),
       PRIMARY_METASTATIC_TUMOR == "Metastatic") %>%
 ggplot(aes(x = .sample, y = counts, fill = PRIMARY_TUMOR_LOCATION)) +
  geom_boxplot()+
  ggtitle("Distribution of Gene Expression per Sample ")+
  xlab("Sample") +
  ylab("Gene Expression (log, CPM)") +
  labs(fill = "Primary Tumor") +
  theme(legend.text = element_text(size = 6))+
  coord_flip() 

ggsave('boxplot_liver_GE.png', width = 8, height = 10)

```

**Fig 2. Looking at gene expression in counts per million to account for size and its distribution for each of the liver metastatic body site samples. Filled by primary tumor location either bile ducts or esophagus. We see ever so slightly esophagus have a little higher median gene expression, is it significant? **


```{r}

ggthemr("light")
plottingMETFilt %>%
  filter(!tumor %in% c("NA","Primary"),
         body == "liver",
         prim_loc %in% c("bile", "esophagus")) %>%
  ggplot(aes(x = prim_loc, y = gene_expression)) +
  geom_boxplot(position = "dodge") +
  labs(x = "Primary Body Site", y = "Total Gene Expression (log/CPM)", color = "Tumor Type") +
  ggtitle("Total Gene Expression Across Primary Tumor Locations")+
  geom_text(data = plottingMETFilt %>% 
              filter(!tumor %in% c("NA","Primary"),body == "liver",prim_loc %in% c("bile", "esophagus")) %>%
              group_by(prim_loc) %>% 
              summarise(top = max(gene_expression), 
                        n=n()),
            aes(x= prim_loc, y=0, label= paste0("n = ", n)), 
            size = 3.5, fontface = "bold", vjust = -2
             )

ggsave('boxplotTOTAL_liver_.png', width = 8, height = 8)

```
**Fig 3. Total gene expression in counts per million (to account for size) and log transformed (for visualization purposes) for all liver metastatic tumors samples. It is comparing the expression between from bile vs esophagus primary tumors and esophagus is slightly higher. They do have comparable sample sizes (31 total samples for this subset).**

```{r}

ggthemr(palette = "light")
sumexp3 %>%
  filter(BODY_SITE == "liver",
      Gender == "M",
       PRIMARY_TUMOR_LOCATION %in% c("bile", "esophagus"),
       PRIMARY_METASTATIC_TUMOR == "Metastatic") %>%
  ggplot(aes(x = counts, color = PRIMARY_TUMOR_LOCATION)) +
  geom_density() +
  xlab("Gene Count (CPM/log)")
  ggtitle("Gene Counts per Body Site") +
  theme(legend.title = element_text(size = 8),
        legend.text = element_text(size = 8))

ggsave('density_plot_Met.png', width = 7, height = 5)
```

#PCA just looking at bile and esophagus prim tumors 
```{r}

#subset just no duplicates abdominal
samp_info_absMET <- sample_info %>%
  filter(BODY_SITE == "liver", 
         Gender == "M",
         PRIMARY_METASTATIC_TUMOR == "Metastatic",
         PRIMARY_TUMOR_LOCATION %in% c("bile", "esophagus")) %>%
  mutate(Year = ifelse(substr(create_date, 1, 4) == "2016", "2016", "2019"))
keep <- samp_info_absMET$Run
sumexpAbs <- sumexp[ ,keep]
sumexpAbs$create_date <- samp_info_absMET$Year

#dds is filtered for this subset question with only primary tumor locations bein bile and esophagus 
assays(sumexpAbs)$counts <- round(assays(sumexpAbs)$counts)
dds2 <- DESeqDataSet(sumexpAbs, design = ~ PRIMARY_TUMOR_LOCATION)
dds2$PRIMARY_TUMOR_LOCATION <- factor(dds2$PRIMARY_TUMOR_LOCATION, levels = c("bile","esophagus"))

vsd2 <- vst(dds2, blind = TRUE)

#method 1
p2 <- pca(assay(vsd2), metadata = colData(vsd2), removeVar = 0.1)
biplot(p2, colby = "PRIMARY_TUMOR_LOCATION", lab = NULL, legendPosition = 'right', legendTitleSize = 8, legendLabSize = 8, title = "PCA of Bile vs Esophagus Tumors" )

#method 2
pcaPlot <- plotPCA(vsd2, intgroup = c("PRIMARY_TUMOR_LOCATION", "create_date", "Age", "LibrarySelection"), returnData = TRUE)
percentVar <- round(100* attr(pcaPlot, "percentVar"))

p <- ggplot(pcaPlot, aes(PC1,PC2)) + 
  geom_point(size = 2, aes(col = LibrarySelection))+
  xlab(paste0("PC1: ", percentVar[1],"% variance"))+
  ylab(paste0("PC2: ", percentVar[2],"% variance"))
p

#ggsave('metpcas.png', width = 8, height = 5)

```
**Fig 4. The PCA shows some distinction between primary tumor locations, however there is some mixing.**
#pca that i use
```{r}
#"liver", "pancreas", 'bile', "prostate", "esophagus"
#just get samples i want and put into sumexperiment
ggthemr("light")
all_prim_loc <- sample_info %>%
  filter(Gender == "M",
         BODY_SITE == "liver",
         PRIMARY_METASTATIC_TUMOR == "Metastatic",
         PRIMARY_TUMOR_LOCATION %in% c("bile", "esophagus")) %>%
  mutate(Year = ifelse(substr(create_date, 1, 4) == "2016", "2016", "2019"),
         Age = cut(Age, breaks = c(0, 19, 39, 59, 79, 93),
                    labels = c("0-19", "20-39", "40-59", "60-79", "80-93")))

all_prim_loc$Year <- factor(all_prim_loc$Year)
all_prim_loc$PRIMARY_TUMOR_LOCATION <- factor(all_prim_loc$PRIMARY_TUMOR_LOCATION,
                                              levels = c("bile","esophagus"))
#which samples to keep in SE
keep <- all_prim_loc$Run
sumexpALL <- sumexp[ , keep]
sumexpALL <- sumexpALL %>%
  mutate(Year = ifelse(substr(create_date, 1, 4) == "2016", "2016", "2019"),
         Age = cut(Age, breaks = c(0, 19, 39, 59, 79, 93),
                    labels = c("0-19", "20-39", "40-59", "60-79", "80-93")))

sumexpALL$Age <- factor(sumexpALL$Age)
#dds object
assays(sumexpALL)$counts <- round(assays(sumexpALL)$counts)
dds3 <- DESeqDataSet(sumexpALL, design = ~ Year + PRIMARY_TUMOR_LOCATION )

#variance stablizing
vsd2 <- vst(dds3, blind = TRUE)

#pcaplot 
pcaPlot <- plotPCA(vsd2, intgroup = c("PRIMARY_TUMOR_LOCATION", "Year", "Age", "LibrarySelection"), returnData = TRUE)
percentVar <- round(100* attr(pcaPlot, "percentVar"))

pcaPlot %>%
  ggplot(aes(PC1,PC2)) + 
  geom_point(size = 3, aes(col = PRIMARY_TUMOR_LOCATION, shape = Year))+
  xlab(paste0("PC1: ", percentVar[1],"% variance"))+
  ylab(paste0("PC2: ", percentVar[2],"% variance")) +
  ggtitle("PCA of Bile vs Esophagus Tumors")+
  theme(legend.title = element_text(size = 7),
        legend.text = element_text(size = 7))

ggsave('Met_pca_new.png', width = 7, height = 5)
```

#pvca plot
```{r}
all_prim_loc <- all_prim_loc %>%
  column_to_rownames("Run")
phenoData <- new("AnnotatedDataFrame", data = all_prim_loc)

eset <- ExpressionSet(assay(vsd2),
                      phenoData = phenoData)
pvcaOBj <- pvcaBatchAssess(eset,batch.factors = c("PRIMARY_TUMOR_LOCATION", "LibrarySelection", "Year", "Age"), 0.6)
ggthemr("dust")

# bp <- barplot(pvcaOBj$dat,
#         ylab = "Weighted average proportion variance",
#          ylim= c(0,1.1),
#         col = c("blue"),
#         las=2,main="PVCA estimation bar chart")
#  axis(1, at = bp, labels = pvcaOBj$label, xlab = "Effects", cex.axis = 0.5, las=2)
#  values = pvcaOBj$dat
#  new_values = round(values , 3)
# text(bp,pvcaOBj$dat,labels = new_values, pos=3, cex = 0.8)

#getting df of pvcaobject 
df <- data.frame(label=as.character(pvcaOBj$label),
                  wmpv=round(as.numeric(pvcaOBj$dat),2))

df$label <- factor(df$label, levels = c("resid", "Year","Age","LibrarySelection","PRIMARY_TUMOR_LOCATION","Year:Age", "LibrarySelection:Age","PRIMARY_TUMOR_LOCATION:Age","PRIMARY_TUMOR_LOCATION:LibrarySelection","LibrarySelection:Year","PRIMARY_TUMOR_LOCATION:Year"))

ggplot(df, aes(x = label, y = wmpv)) +
  geom_bar(stat = "identity") +
  labs(x = "Effects", y = "Weighted average proportion variance",
       title = "PVCA estimation bar chart") +
  scale_y_continuous(limits = c(0, 1.1)) +
  scale_fill_manual(values = "blue") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8),
        axis.text.y = element_text(size = 8),
        legend.position = "none") +
  geom_text(aes(label = round(wmpv, 3), y = round(wmpv, 3)), hjust = -0.2, size = 3.5)+
  coord_flip()

 
ggsave('pvcaMet.png', width = 8, height = 8)
```


```{r}
#DESeq DE
#using medain-of-ratio method to normalize
#estimate size factor for dds
dds2 = estimateSizeFactors(dds3)
#estimating dispersion
dds2 = estimateDispersions(dds2)
#test
dds2 = nbinomWaldTest(dds2)
contrast1 <- c("PRIMARY_TUMOR_LOCATION", "bile", "esophagus")
resDESeq2 = results(dds2, contrast = contrast1)


#LRT
dds2_LRT = DESeq(dds3, test = "LRT", reduced = ~PRIMARY_TUMOR_LOCATION)
contrast1 <- c("PRIMARY_TUMOR_LOCATION", "esophagus","bile")
resDESeq2 = results(dds2_LRT)
#resDESeq2 = results(dds2_LRT,contrast = contrast1,test = "Wald")
r2 = lfcShrink(dds2_LRT,coef = 3, res = resDESeq2)


#ordering by significant p-values
resDESeq2 <- resDESeq2[order(resDESeq2$padj),]
head(resDESeq2)

#results and gene names no filter
sigRes_all <- resDESeq2 %>%
        data.frame() %>%
        rownames_to_column(var = "gene")
sigRes_all <- left_join(sigRes_all, gene_info, by = c("gene" = "GENEID"))
rownames(sigRes_all) <- sigRes_all$gene
sigRes_all <- sigRes_all[,-1]

LFCRes2 <- sigRes_all %>%
        filter(log2FoldChange >= 1 | log2FoldChange <= -1) 
```

```{r}

#select = c("CDCA5","HMGA1", "BRI3BP", "UBE2S", "KIAA0903", "TSPAN5", "FGFR4", "GK5","GASK1A", "ADPRM","MED12L","MYH11","ACTG2", "PRDM8", "MED12L", "ZNF594", "PRDM8")
#select_cool <- c("GK5", "UBE2S", "CDCA5", "BIRC5", "TK1")
select <- c("")
ggthemr("grape")

EnhancedVolcano(toptable = sigRes_all,
                x = "log2FoldChange",
                y = "padj",
                lab = sigRes_all$external_gene_name,
                selectLab = select,
                pCutoff = .05,
                FCcutoff = 1)

ggsave('Met_volcano_plot.png', width = 15, height = 10)


```
**Fig 5. Shows genes in red that both have "significant" differences between the two and have distinct unregulated or downregulated log fold changes (how much gene is expressed).**


```{r}
#getting sample info
DDS2sample_info <- as.data.frame(colData(dds2))

#getting genes
genes.kept <- sigRes4[sigRes4$external_gene_name %in% select_cool, ]
genes.kept <- rownames(genes.kept)

#getting counts/treatment/gene names situated
count.table <- counts(dds2, normalized=T)[genes.kept, ]
cols <- colnames(count.table)
DDS2sample_info <- DDS2sample_info %>%
  filter(rownames(DDS2sample_info) %in% cols)
colnames(count.table) = paste(DDS2sample_info$PRIMARY_TUMOR_LOCATION, colnames(count.table), sep = " : ")

#gene id names
count.table <- count.table %>%
  as.data.frame() %>%
  rownames_to_column(var = "gene")
count.table <-  left_join(count.table, gene_info, by = c("gene" = "GENEID"))
rownames(count.table) <- count.table$external_gene_name
count.table <- count.table[,-33]
count.table <- count.table[,-1]

#heatmap z scores
count.table.z <- t(apply(count.table, 1, scale))
colnames(count.table.z) <- colnames(count.table)

#colors

myCols2 <- c("bile : SRR4307033", "bile : SRR4307364","bile : SRR4306923","bile : SRR4307368","esophagus : SRR8762949","esophagus : SRR8762625","esophagus : SRR8762960","esophagus : SRR8763004")
row_idx2 <- which(colnames(count.table.z) %in% myCols2)


fontcolors <- rep('grey', ncol(count.table.z))
fontcolors[11] <- "black" #esophagus
fontcolors[12] <- "#0000FF" #bile
fontcolors[14] <- "#0000FF"  #bile
fontcolors[15] <- "#0000FF" #bile

fontcolors[23] <- "black" #esophagus
fontcolors[11] <- "#0000FF" #bile
fontcolors[27] <- "black" #esophagus
fontcolors[28] <- "black" #esophagus
fontcolors[29] <- 'black' #esophagus

 fontcolors_rows <- rep('black', nrow(count.table.z))
 fontcolors_rows[1] <- "#0000FF"
 fontcolors_rows[2] <- "#0000FF"
 fontcolors_rows[3] <- "#0000FF"
 fontcolors_rows[5] <- "black"
 fontcolors_rows[4] <- "#0000FF"



#heatmap
h4 <- Heatmap(count.table.z, cluster_rows = T, column_labels = colnames(count.table.z), name = "Z-score", row_labels = rownames(count.table.z),
              column_names_gp = gpar(col = fontcolors, fontface = "bold"),
             row_names_gp = gpar(col = fontcolors_rows, fontface = "bold"))

```

```{r printing heatmap, echo=FALSE}
png("heatmap_MET.png", res = 250, width = 4000, height = 2000)
print(h4)
dev.off()
```
**Fig 6. Tend to be specific unregulated genes in esophagus vs bile. For example CDK5 is upregulated in bile duct cancers and not in esophagus. **


#GSEA
```{r}
#ordered results table
sigResall_gse <- sigRes_all %>%
  filter(!grepl("^ENSG",external_gene_name)) %>%
  rownames_to_column(var = "geneID")

#mading it not be versions
sigResall_gse <- sigResall_gse %>%
  mutate(geneID = str_replace(geneID,"\\..*$", ""))
rownames(sigResall_gse) <- sigResall_gse$geneID
sigResall_gse <- sigResall_gse[,-1]
```

```{r}
#gene list ordered by descending LFC
gene_list2 <- sigResall_gse$log2FoldChange
names(gene_list2) <- rownames(sigResall_gse)
gene <- names(gene_list2)
gene_list2 <- na.omit(gene_list2)
gene_list2 = sort(gene_list2,decreasing = TRUE)
gene_list2
```

```{r}
#running gsea
met_gse <- gseGO(gene_list2,
                    ont = "BP",
                    keyType = "ENSEMBL",
                    OrgDb = "org.Hs.eg.db",
                    eps = 1e-300)

new <- as.data.frame(met_gse)
```

```{r}
gseaplot(met_gse, geneSetID = 102, title = met_gse$Description[102])
ggsave('Met_gsea.png', width = 15, height = 10)
```
**Fig 6. Up regulation of genes in bile in this pathway and down regulated genes in bile in this pathways.** 

```{r}
library(enrichplot)
met_dis <- enricher(gene_list2,
                    org)
barplot(met_gse, showCategory = 10)
```



####NEW HEATMAP and GSEA
```{r}
look_sig <- sigRes_all %>%
 filter(!grepl("^ENSG", external_gene_name),
        padj <= 0.005,
        log2FoldChange <= -1 | log2FoldChange >= 1) %>%
  rownames_to_column(var = "gene") %>%
  mutate(gene = str_replace(gene,"\\..*$", ""))

look_all <- sigRes_all %>%
 filter(!grepl("^ENSG", external_gene_name)) %>%
  rownames_to_column(var = "gene") %>%
  mutate(gene = str_replace(gene,"\\..*$", ""))
```

```{r}
#get list of all hallmark gene sets
library(msigdbr)
hs_msigdb_df <- msigdbr(species = "Homo sapiens")
head(hs_msigdb_df)

#keep only hallmark gene sets
hs_hallmark_df <- hs_msigdb_df %>%
  dplyr::filter(
    gs_cat == "H", # This is to filter only to the H curated gene sets
  )
head(hs_hallmark_df)
```

```{r}
entrez_vector <- mapIds(
  # Replace with annotation package for the organism relevant to your data
  org.Hs.eg.db,
  # The vector of gene identifiers we want to map
  keys = look_all$gene,
  # Replace with the type of gene identifiers in your data
  keytype = "ENSEMBL",
  # Replace with the type of gene identifiers you would like to map to
  column = "ENTREZID",
  # In the case of 1:many mappings, return the
  # first one. This is default behavior!
  multiVals = "first"
)

#get df of ensemble and entrez ids
gene_key_df <- data.frame(
  ensembl_id = names(entrez_vector),
  entrez_id = entrez_vector,
  stringsAsFactors = FALSE
) %>%
  dplyr::filter(!is.na(entrez_id))

#adding gene_key_df to our dataframe of diff expressed genes and universe

universe_df <- gene_key_df %>%
  dplyr::left_join(look_all, by = c("ensembl_id" = "gene"))

diff_df <- look_sig %>%
   dplyr::left_join(gene_key_df, by = c("gene" = "ensembl_id"))%>%
  dplyr::filter(!is.na(entrez_id))
```

```{r}
ORA_results <- enricher(
  gene = diff_df$entrez_id,
  pvalueCutoff = 0.1,
  pAdjustMethod = "BH",
  universe = universe_df$entrez_id,
  TERM2GENE = dplyr::select(
    hs_hallmark_df,
    gs_name,
    human_entrez_gene
  )
)
ora_results <- data.frame(ORA_results@result)

#FDR cutoff
ora_results %>%
  dplyr::filter(p.adjust < 0.1)

barplot(ORA_results, showCategory = 10)

```

```{r}
#list of all genes in the needed gene sets
gene_sets <- ora_results$ID
#totlist of genesets
total_genelist <- hs_hallmark_df %>%
  filter(gs_name %in% gene_sets)

#get list for gsva
hallmarks_list <- split(
  total_genelist$entrez_gene, # The genes we want split into pathways
  total_genelist$gs_name # The pathways made as the higher levels of the list
)

#get vst expression
vst_df <- assay(vsd2) %>%
  as.data.frame() %>% # Make into a data frame
  tibble::rownames_to_column("ensembl_id") %>%
  mutate(ensembl_id = str_replace(ensembl_id,"\\..*$", ""))


mapped_df <- gene_key_df %>%
  dplyr::inner_join(vst_df, by = "ensembl_id")
```

```{r}
# First let's determine the gene means
gene_means <- rowMeans(mapped_df %>% dplyr::select(-ensembl_id, -entrez_id))

# Let's add this as a column in our `mapped_df`.
mapped_df <- mapped_df %>%
  # Add gene_means as a column called gene_means
  dplyr::mutate(gene_means) %>%
  # Reorder the columns so `gene_means` column is upfront
  dplyr::select(ensembl_id, entrez_id, gene_means, dplyr::everything())

filtered_mapped_df <- mapped_df %>%
  # Sort so that the highest mean expression values are at the top
  dplyr::arrange(dplyr::desc(gene_means)) %>%
  # Filter out the duplicated rows using `dplyr::distinct()`
  dplyr::distinct(entrez_id, .keep_all = TRUE)

sum(duplicated(filtered_mapped_df$entrez_id))

filtered_mapped_matrix <- filtered_mapped_df %>%
  # GSVA can't the Ensembl IDs so we should drop this column as well as the means
  dplyr::select(-ensembl_id, -gene_means) %>%
  # We need to store our gene identifiers as row names
  tibble::column_to_rownames("entrez_id") %>%
  # Now we can convert our object into a matrix
  as.matrix()
```

```{r}
gsva_results <- gsva(
  filtered_mapped_matrix,
  hallmarks_list,
  method = "gsva",
  # Appropriate for our vst transformed data
  kcdf = "Gaussian",
  # Minimum gene set size
  min.sz = 15,
  # Maximum gene set size
  max.sz = 500,
  # Compute Gaussian-distributed scores
  mx.diff = TRUE,
  # Don't print out the progress bar
  verbose = FALSE
)
head(gsva_results)
```

```{r}
#body site and primary or met for heatmap
annot_df <- all_prim_loc %>%
  dplyr::select(Run, PRIMARY_TUMOR_LOCATION) 

annot_df <- annot_df[order(annot_df$PRIMARY_TUMOR_LOCATION),]
rownames(annot_df) <- NULL
annot_df <- annot_df %>%
  column_to_rownames("Run")


#get order of columns
names <- rownames(annot_df)

#reorder matrix
gsva_results_order <- gsva_results[,names]
```

```{r}
library(pheatmap)
pheatmap<-pheatmap::pheatmap(
  gsva_results_order,
  annotation_col = annot_df, # Add metadata labels!
  show_colnames = FALSE, # Don't show sample labels
  fontsize_row = 12,
  cluster_col = F,
  main = "Heatmap of Samples and Gene Sets"
)


png("heatmapMET.png", res = 100, width = 2000, height =800)
print(pheatmap)
dev.off()
```






