---
title: "R Notebook"
output: html_notebook
---

```{r libs,eval=TRUE, include=FALSE}
library(tidyverse)
library(DESeq2)
library(purrr)
library(ggplot2)
library(dplyr)
library(RColorBrewer)
library(pheatmap)
library(SummarizedExperiment)
library(knitr)
library(reshape2)
library(limma)
library(edgeR)
library(PCAtools)
library(ggthemr)
library(ComplexHeatmap)
library(EnhancedVolcano)
library(clusterProfiler)
library(org.Hs.eg.db)
library(ggthemr)
library(GSVA)
```


```{r data, eval=TRUE, include=FALSE}
gse <- readRDS(file = "/Users/U1006746/Desktop/counts.AK.output/gene_counts.rds")
list <- c()
```

```{r filter, eval=TRUE, include=FALSE}
#filter out genes (must have more than 2 counts) >= 3???
keep <- rowSums(cpm(gse) > 2) >= 3
se <- gse[keep,]
```

```{r setup, eval=TRUE, include=FALSE}
gene_info <- as.data.frame(rowData(se)@listData)
assay_info <- as.data.frame(assay(se))
sample_info <- as.data.frame(colData(se)@listData)

#how many pairs of tumors/metatstic we have
met_prim_pairs <- sample_info%>%
  dplyr::select(Run,  submitted_subject_id,PRIMARY_METASTATIC_TUMOR, histological_type, body_site) %>%
  filter(PRIMARY_METASTATIC_TUMOR != "NotApplicable") %>%
  group_by(submitted_subject_id) %>%
  filter(any(PRIMARY_METASTATIC_TUMOR == "Primary") & any(PRIMARY_METASTATIC_TUMOR == "Metastatic"))
list <- met_prim_pairs$Run

#filter out columns (pairs)
se<- se[ , list] 
se$PRIMARY_METASTATIC_TUMOR <- factor(se$PRIMARY_METASTATIC_TUMOR, levels = c("Metastatic","Primary"))
se$histological_type <- factor(se$histological_type, levels = c("Not Available","Infiltrating Ductal Carcinoma","Ductal Carcinoma In Situ","Mixture of Tumor and Ductal Carcinoma In Situ","Infiltrating Lobular Carcinoma","Mixed Ductal and Lobular Carcinoma","Invasive Metaplastic Carcinoma with Squamous and Spindle Cell Differentiation", "Invasive Micropapillary Carcinoma","Poorly Differentiated Invasive Ductal Carcinoma" ))

treatment <- colData(se)$PRIMARY_METASTATIC_TUMOR
patient <- colData(se)$submitted_subject_id
body_site <- colData(se)$body_site
run <- colData(se)$Run
table(patient, treatment, body_site, run)


gene_info <- as.data.frame(rowData(se)@listData)
assay_info <- as.data.frame(assay(se))
sample_info <- as.data.frame(colData(se)@listData)
sample_info <- sample_info %>%
  arrange(Run)

# Replace NA values with "."
gene_info$external_gene_name <- ifelse(is.na(gene_info$external_gene_name), gene_info$GENEID, gene_info$external_gene_name)
gene_info$external_gene_name <- ifelse(gene_info$external_gene_name == "", gene_info$GENEID, gene_info$external_gene_name)
gene_info$external_gene_name <- ifelse(gene_info$external_gene_name == "Metazoa_SRP", "Metazoa_SRP.1", gene_info$external_gene_name)



#dds object
se <- se %>%
  mutate(Body = ifelse(BODY_SITE %in% c("breast", "lymph node","brain","liver"), BODY_SITE, "other")) 
#sample_info <- as.data.frame(colData(se)@listData)
assays(se)$counts <- round(assays(se)$counts)
dds <- DESeqDataSet(se, design = ~  submitted_subject_id + PRIMARY_METASTATIC_TUMOR)
dds <- dds[, list]
dds$PRIMARY_METASTATIC_TUMOR <- factor(dds$PRIMARY_METASTATIC_TUMOR, levels = c("Metastatic","Primary"))
dds$submitted_subject_id <- factor(dds$submitted_subject_id)
dds$Body<- factor(dds$Body)
```

```{r transform, eval=TRUE, include=FALSE}
#transforming 
#log transformed
se3 <- se
assay(se3) <- log2(cpm(assay(se3)) + 1)
assaySE3_info <- as.data.frame(assay(se3))
```

**I. Summary Statistics:**

* Breast Cancer Primary tumors from 39 female subjects (no ages given) and matched metastatic tumors
* 29147 genes 
* 102 samples (biopsies)
* 39 subjects 
* Most subjects have 2 samples (1 met 1 primary), one subject has 9 samples (many metastatic sites)
* 63 Metastatic 
* 39 Primary 
* normalized/log transformed/filtered data

```{r}
barplot_data <- data.frame(subject = se3$submitted_subject_id,
                           sample = se3$Run,
                           gene_expression = colSums(assays(se3)$counts),
                           tumor = se3$PRIMARY_METASTATIC_TUMOR)

barplot_data$sample <- factor(barplot_data$sample, levels = unique(barplot_data$sample[order(barplot_data$tumor)]))

boxplot_data <- data.frame(body_site = factor(se3$body_site, 
                                              levels =c("breast","lymph node","brain", "liver","soft tissue",  
                                                        "pleura","chest" ,"lung", "adrenal","bone","diaphragm","ovary","pericardium","peritoneum","rectum","skin","spleen","stomach","thyroid","uterus")),
                                              gene_expression = colSums(assays(se3)$counts))
boxplot_data <- boxplot_data %>%
  rownames_to_column(var = "sample")

```

```{r}
library(tidySummarizedExperiment)
se3 <- se3 %>%
  group_by(PRIMARY_METASTATIC_TUMOR)
se3$.sample <- factor(se3$.sample, levels = unique(se3$.sample[order(se3$PRIMARY_METASTATIC_TUMOR)]))
se3 <- se3[, -c(7:47)]


se3 <- se3 %>%
  mutate(Body = ifelse(BODY_SITE %in% c("breast", "lymph node","brain","liver"), BODY_SITE, "other")) 

```

```{r}

ggthemr("grape")
se3 %>%
 ggplot(aes(x = .sample, y = counts, fill = PRIMARY_METASTATIC_TUMOR)) +
  geom_boxplot()+
  ggtitle("Distribution of Gene Expression per Sample ")+
  xlab("Sample") +
  ylab("Gene Expression (log, CPM)") +
  labs(fill = "Tumor Type") +
  theme(axis.text.x = element_text(angle = 90),
         legend.title = element_text(size = 15),
        legend.text = element_text(size = 15),
        plot.title = element_text(size = 20))
  

ggsave('boxplot_gene_exp.png', width = 25, height = 15)
```

*Fig 1. The distribution of gene expression for every sample. There are no major outliers and median tends to be slightly higher in primary tumors as compared to metastatic tumors. These genes are normalized to account for gene length and are log transformed for better interpretation.*

```{r}


ggthemr(palette = "light")
se3 %>%
  ggplot(aes(x = counts, color = Body)) +
  geom_density() +
  ggtitle("Gene Counts per Body Site") +
  xlab("Gene Count (CPM/log)")
  theme(legend.title = element_text(size = 8),
        legend.text = element_text(size = 8))

ggsave('density_plot_AUR.png', width = 7, height = 5)
```

*Fig 1.b Looking at gene counts for specific body type. *


```{r}

boxplot_data$body_site <- factor(boxplot_data$body_site)
ggthemr("grape")
boxplot_data %>%
  ggplot(aes(x = body_site, y = gene_expression, fill = ifelse(body_site == "breast", "breast", "non-breast"))) +
  geom_boxplot(position = "dodge") +
  labs(x = "Tumor Site", y = "Total Gene Expression (log/CPM)", color = "Tumor Type") +
  ggtitle("Total Gene Expression Across Body Sites")+
  theme(axis.text.x = element_text(angle = 90, face = c("plain","bold", "bold", "bold","plain","plain","plain","plain","plain","plain","plain","plain","plain","plain","plain","plain","plain","plain","plain","plain" )),
        legend.position = "none")+
  geom_text(data = boxplot_data %>% 
              group_by(body_site) %>% 
              summarise(top = max(gene_expression), 
                        n=n()),
            aes(x= body_site, y=0, label= paste0("n = ", n)), 
            size = 2.8, fontface = "bold", vjust = -2)

ggsave('boxplot_body.png', width = 8, height = 8)

```
**Fig 2. Boxplot of each tumor site with primary location in red and the relative total gene expression (in counts per million log transformed to account for size better visualization) for every sample within that body site. Number of samples per body site are listed below the respective boxplot. As you see breast as it is the primary location has 39 samples (one for every subject) and has the overall largest gene expression totals for those samples. Lymph nodes, brain and liver metastatic sites all have very similar sample sizes and distributions. With samples of the brain having slightly higher total gene expression compared to other sites. Most of the other body sites only have 1-2 samples and are not very informative.**

```{r}
ggthemr("light")
boxplot_data %>%
  filter(body_site != "breast") %>%
  ggplot(aes(x = body_site, fill = ifelse(body_site %in% c("lymph node", "brain", "liver"), "brain", "non-breast"))) +
  geom_bar() +
  scale_y_continuous(labels = scales::comma, breaks = seq(0, 12, by = 2))+
  theme(axis.text.x = element_text(angle = 90)) +
  ggtitle("Number Samples Per Body Site (Metastatic Tumors Only)")+
  xlab("Metastatic Body Site") +
  ylab("Number of Samples")+
  theme(legend.position = "none")

ggsave('barplot_body_smaples.png', width = 8, height = 8)
```
**Fig 3. A visual to show the number of samples with in each body site.  In blue are the body sites with highest number of samples, lymph nodes, brain and liver. Plus 39 samples for breast since it is the primary for every sample.**


```{r}
ggthemr("light")
barplot_data %>%
  ggplot(aes(x = subject, fill = tumor)) +
  geom_bar() +
  ggtitle("Number Samples per Subject")+
  xlab("Subject") +
  ylab("Number of Samples") +
  theme(
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 10),
    axis.text.x = element_text(angle = 90, margin = margin(t = 5)),
    axis.title.x = element_text(margin = margin(t = 10, b = 20)),
    axis.title.y = element_text(margin = margin( r = 10)),
    plot.title = element_text(hjust = 0.5),
    plot.caption.position = "panel" ,
    plot.caption = element_text(margin = margin(l = 60, r = 60)),
    plot.margin = unit(c(0.25, 0.2, 0.2, 0.25),"inches"))+
  scale_x_discrete(labels=c("AUR-AER6"=expression(bold("AUR-AER6")), "AUR-AFEA"=expression(bold("AUR-AFEA")),
                              "AUR-AFE9"=expression(bold("AUR-AFE9")), parse=TRUE)) +
    scale_y_continuous(labels = scales::comma, breaks = seq(0, 10, by = 2))
ggsave('num_sams_aurora.png', width = 12, height = 12)
```
**Fig. 3b. The number of metastatic samples for each of the 39 subjects. In bold are the subjects with highest number of samples.**

```{r}
ggthemr("light")
vsd <- vst(dds, blind = TRUE)

#pca 1
# p <- pca(assay(vsd), metadata = colData(vsd), removeVar = 0.1)
# biplot(p, colby = "PRIMARY_METASTATIC_TUMOR", lab = NULL, legendPosition = 'right', legendTitleSize = 8, legendLabSize = 8, title = "PCA of Primary vs Metastatic Tumors" )


#pca 2
pcaPlot <- plotPCA(vsd, intgroup = c("PRIMARY_METASTATIC_TUMOR", "submitted_subject_id", "Body"), returnData = TRUE)
percentVar <- round(100* attr(pcaPlot, "percentVar"))
highlighted_points <- pcaPlot %>%
  filter(PRIMARY_METASTATIC_TUMOR == "Metastatic" & PC1 < 0 & PC2 < -15)

library(ggthemr)
ggthemr("dust")
ggplot(pcaPlot, aes(PC1,PC2)) + 
  geom_point(size = 2, aes(col = PRIMARY_METASTATIC_TUMOR, shape = Body), show.legend = TRUE) +
  stat_ellipse(data = subset(pcaPlot, Body != "breast"), aes(fill = Body), show.legend = FALSE)+
  xlab(paste0("PC1: ", percentVar[1],"% variance"))+
  ylab(paste0("PC2: ", percentVar[2],"% variance"))+
  ggtitle("PCA of Primary vs Metastatic Tumors")+
  geom_segment(data = highlighted_points,
               aes(x = PC1, xend = PC1, y = PC2, yend = PC2 - 20),  
               arrow = arrow(type = "closed", length = unit(.05, "inches")),
               linejoin = "mitre", color = "black") +
  geom_text(data = highlighted_points,
            aes(label = Body, x = PC1, y = PC2 - 3), 
            vjust = 10, fontface = "bold") +
  theme_classic() +
   geom_text(data = data.frame(label = c("liver (n = 12)", "lymph node (n = 9)", "other (n = 33)", "brain (n = 9)"),x = c(50, -25, -25, 60), y = c(70, 65, 50, -40)),
            aes(x = x, y = y, label = label), vjust = 2, hjust = 0, fontface= "bold")+
  guides(col = guide_legend("Tumor Type"), fill = "none",shape = guide_legend(override.aes = list(col = "black")))
              
ggsave('apcas.png', width = 10, height = 8)

```

```{r}
pca_result <- prcomp(t(assay(vsd)), scale. = TRUE)

# Get the rotated component scores
rotated_scores <- pca_result$x
sdev <- pca_result$sdev

# Calculate the percentage of variation explained by each PC
variance_explained <- (sdev^2) / sum(sdev^2) * 100

# Create a data frame
pca_scores <- data.frame(PC3 = rotated_scores[, 3],
                         PC4 = rotated_scores[, 4],
                         PRIMARY_METASTATIC_TUMOR = colData(vsd)$PRIMARY_METASTATIC_TUMOR)

# Create a scatter plot of PC3 vs PC4
ggplot(pca_scores, aes(x = PC3, y = PC4, color = PRIMARY_METASTATIC_TUMOR)) +
  geom_point() +
  labs(title = "PCA of Bile vs Esophagus Tumors (PC3 vs PC4)",
       x = "PC3", y = "PC4") +
   xlab(paste0("PC3: ", variance_explained[3],"% variance"))+
  ylab(paste0("PC4: ", variance_explained[4],"% variance"))+
  theme_minimal()

```

**Fig 4. PCA shows there are separate groups between metastatic and primary tumors.**

#Differential Gene Expression
```{r diff, eval=TRUE, include=FALSE}
############wald based
#using medain-of-ratio method to normalize
#estimate size factor for dds
dds = estimateSizeFactors(dds)
# #estimating dispersion
dds = estimateDispersions(dds)
# #test
dds = nbinomWaldTest(dds)
##results and contrast
contrast <- c("PRIMARY_METASTATIC_TUMOR", "Metastatic", "Primary")
resDESeq = results(dds, contrast = contrast)

contrast2 <- c("Body", "liver", "lymph node")
resDESeq = results(dds, contrast = contrast2)

############LRT
dds_LRT = DESeq(dds, test = "LRT", reduced = ~PRIMARY_METASTATIC_TUMOR)
resDESeq = results(dds_LRT)
resDESeq <- resDESeq[order(resDESeq$padj),]
head(resDESeq)
#summary of <0.05 p-value
summary(resDESeq, alpha = 0.05)

#just names in full dataset
resDESeq_names <- resDESeq %>%
        data.frame() %>%
        rownames_to_column(var = "gene")
resDESeq_names <- left_join(resDESeq_names, gene_info, by = c("gene" = "GENEID"))
rownames(resDESeq_names) <- resDESeq_names$gene
resDESeq_names <- resDESeq_names[,-1]


#just sig genes
padj.cutoff <- 0.05
sigRes <- resDESeq %>%
        data.frame() %>%
        filter(padj < padj.cutoff) %>%
        rownames_to_column(var = "gene")

LFCRes <- resDESeq_names %>%
        filter(log2FoldChange >= 1 | log2FoldChange <= -1) 

#gene symbols in results table 
sigRes <- left_join(sigRes, gene_info, by = c("gene" = "GENEID"))
rownames(sigRes) <- sigRes$gene
sigRes <- sigRes[,-1]

head(sigRes)
```
#GSEA
#gene list is look_sig these are DEG
```{r}
look_sig <- resDESeq_names %>%
 filter(!grepl("^ENSG", external_gene_name),
        padj <= 0.005,
        log2FoldChange <= -1 | log2FoldChange >= 1) %>%
  rownames_to_column(var = "gene") %>%
  mutate(gene = str_replace(gene,"\\..*$", ""))

look_all <- resDESeq_names %>%
 filter(!grepl("^ENSG", external_gene_name)) %>%
  rownames_to_column(var = "gene") %>%
  mutate(gene = str_replace(gene,"\\..*$", ""))
```

```{r}
#get list of all hallmark gene sets
library(msigdbr)
hs_msigdb_df <- msigdbr(species = "Homo sapiens")
head(hs_msigdb_df)

#keep only hallmark gene sets
hs_hallmark_df <- hs_msigdb_df %>%
  dplyr::filter(
    gs_cat == "H", # This is to filter only to the H curated gene sets
  )
head(hs_hallmark_df)
```

```{r}
entrez_vector <- mapIds(
  # Replace with annotation package for the organism relevant to your data
  org.Hs.eg.db,
  # The vector of gene identifiers we want to map
  keys = look_all$gene,
  # Replace with the type of gene identifiers in your data
  keytype = "ENSEMBL",
  # Replace with the type of gene identifiers you would like to map to
  column = "ENTREZID",
  # In the case of 1:many mappings, return the
  # first one. This is default behavior!
  multiVals = "first"
)

#get df of ensemble and entrez ids
gene_key_df <- data.frame(
  ensembl_id = names(entrez_vector),
  entrez_id = entrez_vector,
  stringsAsFactors = FALSE
) %>%
  dplyr::filter(!is.na(entrez_id))

#adding gene_key_df to our dataframe of diff expressed genes and universe

universe_df <- gene_key_df %>%
  dplyr::left_join(look_all, by = c("ensembl_id" = "gene"))

diff_df <- look_sig %>%
   dplyr::left_join(gene_key_df, by = c("gene" = "ensembl_id"))%>%
  dplyr::filter(!is.na(entrez_id))
```

```{r}
ORA_results <- enricher(
  gene = diff_df$entrez_id,
  pvalueCutoff = 0.1,
  pAdjustMethod = "BH",
  universe = universe_df$entrez_id,
  TERM2GENE = dplyr::select(
    hs_hallmark_df,
    gs_name,
    human_entrez_gene
  )
)
ora_results <- data.frame(ORA_results@result)

#FDR cutoff
ora_results %>%
  dplyr::filter(p.adjust < 0.1)

barplot(ORA_results, showCategory = 10)

```


#GSVA
```{r}
#list of all genes in the needed gene sets
gene_sets <- ora_results$ID
#totlist of genesets
total_genelist <- hs_hallmark_df %>%
  filter(gs_name %in% gene_sets)

#get list for gsva
hallmarks_list <- split(
  total_genelist$entrez_gene, # The genes we want split into pathways
  total_genelist$gs_name # The pathways made as the higher levels of the list
)

#get vst expression
vst_df <- assay(vsd) %>%
  as.data.frame() %>% # Make into a data frame
  tibble::rownames_to_column("ensembl_id") %>%
  mutate(ensembl_id = str_replace(ensembl_id,"\\..*$", ""))


mapped_df <- gene_key_df %>%
  dplyr::inner_join(vst_df, by = "ensembl_id")
```

```{r}
# First let's determine the gene means
gene_means <- rowMeans(mapped_df %>% dplyr::select(-ensembl_id, -entrez_id))

# Let's add this as a column in our `mapped_df`.
mapped_df <- mapped_df %>%
  # Add gene_means as a column called gene_means
  dplyr::mutate(gene_means) %>%
  # Reorder the columns so `gene_means` column is upfront
  dplyr::select(ensembl_id, entrez_id, gene_means, dplyr::everything())

filtered_mapped_df <- mapped_df %>%
  # Sort so that the highest mean expression values are at the top
  dplyr::arrange(dplyr::desc(gene_means)) %>%
  # Filter out the duplicated rows using `dplyr::distinct()`
  dplyr::distinct(entrez_id, .keep_all = TRUE)

sum(duplicated(filtered_mapped_df$entrez_id))

filtered_mapped_matrix <- filtered_mapped_df %>%
  # GSVA can't the Ensembl IDs so we should drop this column as well as the means
  dplyr::select(-ensembl_id, -gene_means) %>%
  # We need to store our gene identifiers as row names
  tibble::column_to_rownames("entrez_id") %>%
  # Now we can convert our object into a matrix
  as.matrix()
```

```{r}
gsva_results <- gsva(
  filtered_mapped_matrix,
  hallmarks_list,
  method = "gsva",
  # Appropriate for our vst transformed data
  kcdf = "Gaussian",
  # Minimum gene set size
  min.sz = 15,
  # Maximum gene set size
  max.sz = 500,
  # Compute Gaussian-distributed scores
  mx.diff = TRUE,
  # Don't print out the progress bar
  verbose = FALSE
)
head(gsva_results)
```

```{r}
#body site and primary or met for heatmap
annot_df <- sample_info %>%
  dplyr::select(Run, PRIMARY_METASTATIC_TUMOR, Body) %>%
  column_to_rownames("Run")

annot_df <- annot_df[order(annot_df$PRIMARY_METASTATIC_TUMOR), ]

#for liver and lymph
# annot_df <- annot_df %>%
#   select(Body) %>%
#   filter(Body %in% c("liver", "lymph node")) %>%
#   arrange(Body)

#get order of columns
names <- rownames(annot_df)

#reorder matrix
gsva_results_order <- gsva_results[,names]
```

```{r}

pheatmap<-pheatmap::pheatmap(
  gsva_results_order,
  annotation_col = annot_df, # Add metadata labels!
  show_colnames = FALSE, # Don't show sample labels
  fontsize_row = 10,
  cluster_col = F,
  main = "Heatmap of Samples and Gene Sets"
)

#ggsave('gsva_heatmap.png', width = 15, height = 10)

png("heatmap.png", res = 100, width = 2000, height = 1000)
print(pheatmap)
dev.off()
```


```{r}
select4 = c("")

ggthemr("grape")
EnhancedVolcano(toptable = resDESeq_names,
                x = "log2FoldChange",
                y = "padj",
                FCcutoff= 1,
                pCutoff = 0.005,
                lab = resDESeq_names$external_gene_name,
                selectLab = select4)




ggsave('volcano_plot.png', width = 15, height = 10)
```
**Fig 5. Volcano plot shows p-value significance on y (0.05) and log fold change (either up or down regulated depending on positive or negative) on x. We are interested in genes high above dashed line and far from center line. **



#didnt use theses next two figures
```{r heatmap, echo=FALSE}
#getting sample info
DDSsample_info <- as.data.frame(colData(dds))

#getting genes
genes.kept <- sigRes[sigRes$external_gene_name %in% select4, ]
genes.kept <- rownames(genes.kept)

#getting counts/treatment/gene names situated
count.table <- counts(dds, normalized=T)[genes.kept, ]
cols <- colnames(count.table)
DDSsample_info <- DDSsample_info %>%
  filter(rownames(DDSsample_info) %in% cols)
colnames(count.table) = paste(DDSsample_info$PRIMARY_METASTATIC_TUMOR, colnames(count.table), sep = " : ")

#gene id names
count.table <- count.table %>%
  as.data.frame() %>%
  rownames_to_column(var = "gene")
count.table <-  left_join(count.table, gene_info, by = c("gene" = "GENEID"))
rownames(count.table) <- count.table$external_gene_name
count.table <- count.table[,-104]
count.table <- count.table[,-1]

col_sams <- colnames(count.table)
rows <- rownames(count.table)

myCols <- c("Metastatic : SRR21068776", "Metastatic : SRR21068781","Metastatic : SRR21068682","Metastatic : SRR21068772","Primary : SRR21068765","Primary : SRR21068641","Primary : SRR21068764","Primary : SRR21068782")
#row_idx <- which(colnames(count.table.z) %in% myCols)


fontcolors <- rep('grey', ncol(count.table.z))
fontcolors[35] <- 'black'
fontcolors[53] <- "black"
fontcolors[57] <- "black"
fontcolors[49] <- 'black'
fontcolors[43] <- "#0000FF"
fontcolors[6] <- "#0000FF"
fontcolors[42] <- "#0000FF"
fontcolors[58] <- "#0000FF"

fontcolors_rows <- rep('black', nrow(count.table.z))
fontcolors_rows[1] <- "black" 
fontcolors_rows[2] <- "black"
fontcolors_rows[3] <- "#0000FF"
fontcolors_rows[4] <- "#0000FF"

#heatmap z scores
count.table.z <- t(apply(count.table, 1, scale))
colnames(count.table.z) <- colnames(count.table)
#count.table.z.filtered <- count.table.z[, apply(count.table.z, 2, function(x) all(x > threshold_value))]

#heatmap
h2 <- Heatmap(count.table.z, 
              cluster_rows = T, 
              column_labels = col_sams, 
              name = "Z-score",
              row_labels = rownames(count.table.z),
              column_names_gp = gpar(col = fontcolors, fontface = "bold",
                                     fontsize = "14"),
              row_names_gp = gpar(col = fontcolors_rows, fontface = "bold", 
                                  fontsize = "15")
              )
```

```{r printing heatmap, echo=FALSE}
png("heatmap.png", res = 100, width = 2800, height = 1500)
print(h2)
dev.off()
```
**Fig 6. This heat map is a visual representation of hierarchical cluster the samples with corresponding tumor types. We see that most metastatic tumors have more up regulated genes in the significant genes we picked as compared to primary tumors. Also see there is a smattering of metastatic that are not unregulated and more closley resemble the primary tumors.**

```{r}
#ordered results table

resDESeq_gse <- resDESeq_names %>%
  filter(!grepl("^ENSG",external_gene_name)) %>%
  rownames_to_column(var = "geneID")
resDESeq_gse <- resDESeq_gse %>%
  mutate(geneID = str_replace(geneID,"\\..*$", ""))
rownames(resDESeq_gse) <- resDESeq_gse$geneID
resDESeq_gse <- resDESeq_gse[,-1]
```

```{r}
gene_list <- resDESeq_gse$log2FoldChange
names(gene_list) <- rownames(resDESeq_gse)
gene_list<- na.omit(gene_list)
gene_list = sort(gene_list,decreasing = TRUE)
gene_list
```

```{r}
aurora_gse <- gseGO(gene_list,
                    ont = "BP",
                    keyType = "ENSEMBL",
                    OrgDb = "org.Hs.eg.db",
                    eps = 1e-300)

as.data.frame(aurora_gse)

```

```{r}
gseaplot(aurora_gse, geneSetID = 4, title = aurora_gse$Description[4])
ggsave('Aurora_gsea.png', width = 15, height = 10)
```

**Fig 7. The gene set of antigen receptor-mediated signaling pathway is down regulated in metastatic to primary, meaning that this could cause more issues for the immune system to fight and recognize the tumor and contribute to the progression of the metastatic tumors. Most of the genes in the geneset have negative log fold changes at end of ranked list. RSE is cumulative.**
